---
title: "Getting started with micsPlusTableR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with micsPlusTableR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`micsPlusTableR` turns an Excel tabulation plan, survey data, and a preparation
script into checked survey tables. Use its Shiny application for interactive
review or its R functions for a scripted workflow.

## Set up your survey project first

For a beginner walkthrough, read the
[User's Guide (PDF)](https://github.com/bariscr/micsPlusTableR/blob/HEAD/inst/doc/user-guide.pdf).
It covers installing R and RStudio, project creation, package installation,
file selection, and the application's buttons with screenshots.

1. Install [R](https://cloud.r-project.org/) (4.2 or later), then
   [RStudio Desktop](https://posit.co/download/rstudio-desktop/).
2. Create your own survey project using **File > New Project > New Directory
   > New Project**. For an existing survey folder without an `.Rproj` file,
   choose **Existing Directory**. For an existing RStudio project, use
   **File > Open Project** and select its `.Rproj` file.
3. Confirm the survey project name at the top right. In the **R Console**, run
   the installation commands once, waiting for each to finish:

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("bariscr/micsPlusTableR")
```

4. Click **Session > Restart R**. Check `packageVersion("micsPlusTableR")`, then
   launch the app as shown below and select the survey inputs.

You can also use **Positron**, **VS Code**, **Cursor**, **JupyterLab/Notebook with
an R kernel**, or **R's console/a terminal**. Configure R support, open your
survey folder, and start an R session there. Run commands in R, not in a Python
kernel or an ordinary shell. Check `getwd()` and `here::here()` for the intended
folder, or pass `output_dir` explicitly. Use `launch.browser = TRUE` if needed.
Remote sessions require appropriate browser routing and access to server files.
The User's Guide explains the alternatives and links to their setup instructions.

On subsequent visits, reopen your survey project and launch the app; select
inputs again for the new session. Each user's project holds their survey work.
The package source and maintainer's management project are not needed to use
an installed package.

## Download tabulation plans and preparation files

Tabulation plans, preparation scripts, and supporting reference assets are maintained in
[micsPlusTableR-Files](https://github.com/bariscr/micsPlusTableR-Files).
Household and household-member microdata are **not** distributed there or by
these functions. Obtain the `.sav` data files separately from the
[MICS surveys page](https://mics.unicef.org/surveys) with the required permission.
Once the files repository is public, no GitHub account is needed for its
plans and preparation files. You can
[download all plans and preparation files as a ZIP](https://github.com/bariscr/micsPlusTableR-Files/archive/refs/heads/main.zip)
and extract it, use the app tab, or run the Console commands below.

### Download from the application

The app opens on **Data Preparation**. To download files, open **Download Files** and click
**Load / refresh surveys**, then choose **Selected surveys** and tick one or
more folders, or choose **All surveys**. The preview shows the selected file
count and size. **Download folder** defaults to the current project captured
when `run_app()` starts. Each survey is saved in its own subfolder, such as
`<project>/JAM_W1/`.

Use **Browse folders...** or type another path; relative paths are resolved
against the current project. **Use current project** restores the default.
Leave **Replace existing files** off unless you intend to replace local copies.
Click **Download files**, wait for the saved-location message, then open
**Data Preparation** and select the materials together with your separately
authorized household/member data. On a remote R host, these paths refer to
that host's filesystem.

### Download from the R Console

The functions below are an alternative to the app tab. Their default destination
is `inputs/` under `getwd()`; the app tab instead defaults to the project itself.

```{r survey-downloads, eval=FALSE}
# See available survey IDs and file paths
available <- micsPlusTableR::list_survey_files()
unique(available$survey)

# Plans and complete preparation folder for one survey
micsPlusTableR::download_survey_files("JAM_W1")

# Several surveys
micsPlusTableR::download_survey_files(c("JAM_W1", "MNG_W2"))

# All published surveys
micsPlusTableR::download_survey_files()

# Individual files: use exact paths from available$path
micsPlusTableR::download_survey_files(
  files = available$path[1],
  dest_dir = "selected-inputs"
)
```

These are alternative examples. Downloads go under `inputs` in the current
working directory (`getwd()`), keeping folders such as `JAM_W1`
and all preparation subfolders. To replace an existing selection deliberately,
pass `overwrite = TRUE`; otherwise an existing file stops the download.
Select a whole survey folder to include all its preparation helpers. These
downloads do not include the household/member data needed to run the app.
Save your separately authorized data alongside the matching plans and scripts.
Downloaded scripts are saved without being executed. Select the files in the app as usual.
New survey folders are discovered from GitHub without updating the package.
Use `ref = "<commit-or-tag>"` to retrieve a recorded version; the returned table's
`commit` attribute records the resolved commit. Internet access and a public
files repository are required. If GitHub limits requests, retry later or use
the ZIP link. Before publication, obtain the plans and preparation files from
your survey lead.

## Inputs

Prepare these files before starting:

1. An Excel plan with an `IDX` sheet and one worksheet per table.
2. Household (`hh`) and household-member (`hl`) SPSS `.sav` files obtained
   separately from the MICS Plus website with permission.
3. A preparation folder with one top-level prep-like `.R` script. The script
   reads `hh_path` and `hl_path` and creates data frames named `hh` and `hl`.
   Optional FIES assets belong in a sibling `FIES-inputs` subdirectory.

Preparation scripts execute R code and define survey-specific recodes and
additional tables. Review and validate them for the survey being analysed.
Plans and preparation files are downloaded separately from `micsPlusTableR-Files`.
Microdata are obtained from the MICS Plus website with permission. Neither is
bundled in the package.

## Interactive use

```{r launch, eval=FALSE}
library(micsPlusTableR)
run_app()
# Or select the output location:
run_app(output_dir = "survey-results")
```

Select the survey and wave, provide the plan, data files, and preparation
folder, then prepare the data. Calculate the required tables, inspect the
consistency checks, and export the workbooks. By default, workbooks are written
to `micsPlusTableR-output` below the project root detected by `here::here()`.

The **Long Format Data** tab offers Excel, CSV, and RDS downloads. Downloads
include the full prepared dataset even when the preview is filtered. RDS
preserves R column types. The SDMX-compatible fields follow the package's
observation contract; review requirements of any external SDMX system separately.

## Scripted use

A session keeps prepared data, the table plan, and results together. Create a
separate session for independent workflows:

```{r session}
library(micsPlusTableR)
session <- mics_session()
session
```

The following example requires your own input files:

```{r workflow, eval=FALSE}
prepare_mics_data(
  session,
  hh_path = "hh.sav",
  hl_path = "hl.sav",
  prep_script = "survey-preparation/survey_prep.R",
  output_dir = "survey-results"
)
result <- tabulate_mics_table(session, "TabulationPlan.xlsx", "1.1")
checks <- check_mics_table(session, result)
preview <- pivot_mics_table(session, result, type = "header")

workbooks <- create_excel_workbooks(
  "TabulationPlan.xlsx", "survey-results",
  country_code = "JAM", period = "2023-24", wave = "Wave 1"
)
write_mics_table(session, workbooks$output, sheet = "1.1", table = result)
write_mics_footnotes(session, workbooks$output, sheet = "1.1", table = result)
```

`compare_mics_tables()` and `read_previous_mics_table()` support review against
previous results. Consistency checks help find issues but cannot validate the
substantive choices in a preparation script or tabulation plan.

## Test individual steps in R

The smaller engine helpers are available directly from the package. Stateful
helpers take `session` as their first argument, so independent sessions do not
share results. The direction-specific names are now `tabulate_v()` and
`tabulate_h()`; replace any old `tabulate_v2()` or `tabulate_h2()` calls. Calls
through `tabulate_mics_table()` continue to select the direction automatically.

```{r individual-checks, eval=FALSE}
# After preparing session and reading a plan:
plan <- read_mics_tabulation(session, "TabulationPlan.xlsx", "1.1")
row_rules <- row_condition_f(plan$tab_r)
column_rules <- col_condition_f(plan$tab_c)
result <- if (plan$tab_direction == "v") tabulate_v(session) else tabulate_h(session)

# Run only the check of interest, using long-format numeric results:
row_check <- row_group_total_check(session, result, diff = 1e-6)
row_check$status
row_check$issues
```

Each individual check returns `status`, `issue_count`, `details` (all
comparisons), and `issues` (failing comparisons). `not_applicable` means the
required total or distribution is absent; it is not a passing validation.
`check_mics_table()` still runs all six checks with a default tolerance of
`1e-6`; individual helpers retain the engine default `diff = 1e-11`.

| Function | Purpose |
|---|---|
| `tabulate_v()`, `tabulate_h()` | Calculate the currently loaded vertical or horizontal plan and save results in the session. |
| `tabulate_extra_table()` | Map a supplied numeric table to the plan by position, retaining row logic as labels. |
| `calc_cells()` | Test individual horizontal cells using explicit data, row/column specifications, statistics, and weights. |
| `row_condition_f()`, `col_condition_f()` | Inspect cleaned predicates and generated variable names; row output also separates calculations. |
| `normalize_condition_text()` | Convert Excel smart quotes and Unicode operators to R syntax. |
| `extract_var()` | Extract the leading variable from a supported condition; return NA for unsupported syntax. |
| `get_blank_cols()` | Identify blank header columns in worksheet rows 3–8. |
| `filter_second()` | Remove rows whose second column is NA. |
| `row_group_total_check()` | Compare row-group counts with the overall total. |
| `row_group_perc_total_check()` | Check row-group percentage totals. |
| `col_group_total_check()` | Compare column-group counts with the overall total. |
| `col_group_perc_total_check()` | Check column-block percentages against 100. |
| `row_indent_group_total_check()` | Compare indented child counts with their parent row. |
| `row_indent_group_perc_total_check()` | Compare indented child percentages with their parent row. |

A complete, runnable cell example:

```{r individual-cell}
calc_cells(
  df = data.frame(total = 1, sex = c(1, 2, 2), weight = c(1, 2, 3)),
  tab_r = data.frame(row_index = 9L, row_lgc = "TRUE"),
  tab_c3 = data.frame(col_index = 3L, col_condition = "sex == 2",
                       col_var_name = "sex_2"),
  tab = data.frame(row_index = 9L, col_index = 3L, stat_type = "n"),
  weight_var = "weight", weighted = TRUE
)
```

This returns a weighted count of 5. `weighted = FALSE` returns 2. For historical
compatibility, `weighted = "both"` returns the unweighted result only; call the
helper twice to compare both modes. Use `?calc_cells` for supported statistics
and their weighting rules. Direction helpers return the engine result;
`tabulate_mics_table()` additionally attaches variable-explanation labels.

Errors identify the worksheet, block or cell and the relevant expression when
available. A missing-variable message means the condition or statistic refers
to a column absent from that block's prepared data. Fix the named input and
rerun. Use raw long-format results for checks, not pivoted or formatted tables.
Each helper's help page documents arguments, return values, and requirements.

## Edit surveys and waves (maintainers)

In the development source, edit **`inst/extdata/survey_choices.csv`**. It is a
normal CSV and remains editable when stored on GitHub. R includes it in the
installed package automatically. Its current contents can be read with:

```{r survey-choices}
path <- system.file("extdata", "survey_choices.csv", package = "micsPlusTableR")
read.csv(path, colClasses = "character", fileEncoding = "UTF-8-BOM")
```

Keep the four columns `country`, `country_code`, `period`, and `wave`. Add one
row per available combination, with no blank fields or duplicates. Country
codes must be three uppercase letters and consistent within a country/period.
Treat periods such as `2025-26` as text in spreadsheet editors. Save as UTF-8.
Row order controls dropdown order; the first survey and its first wave are the
defaults. Only the selected survey's waves are offered.

From the package root, `source("run_dev.R")` loads the source files directly.
Stop the app and relaunch after saving changes. For an installed package,
reinstall the updated source and restart. Avoid maintaining the installed CSV:
installation replaces that copy.

## Documentation

Use `help(package = "micsPlusTableR")` for the function index and
`help("survey_choices", package = "micsPlusTableR")` for the CSV specification.
The detailed application guide is available at:

```{r guide}
system.file("doc", "user-guide.pdf", package = "micsPlusTableR")
```

## Package distribution (maintainers)

The repository root is already a standard R source package. Commit the source
files to GitHub, including the CSV, `man/` help pages, and this vignette. From
that root, `Rscript tools/build-package.R` validates the CSV and creates an
installable source archive in `build/`. `Rscript tools/build-manual.R` creates
the CRAN-style reference PDF from the help pages in `output/pdf/`.

No conversion or duplicate package folder is needed before pushing to GitHub.
