---
title: "Getting started with micsPlusTableR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with micsPlusTableR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`micsPlusTableR` turns an Excel tabulation plan, survey data, and a preparation
script into checked survey tables. Use its Shiny application for interactive
review or its R functions for a scripted workflow.

## Inputs

Prepare these files before starting:

1. An Excel plan with an `IDX` sheet and one worksheet per table.
2. Household (`hh`) and household-member (`hl`) SPSS `.sav` files.
3. A preparation folder with one top-level prep-like `.R` script. The script
   reads `hh_path` and `hl_path` and creates data frames named `hh` and `hl`.
   Optional FIES assets belong in a sibling `FIES-inputs` subdirectory.

Preparation scripts execute R code and define survey-specific recodes and
additional tables. Review and validate them for the survey being analysed.
Survey input files are supplied by the analyst and are not bundled here.

## Interactive use

```{r launch, eval=FALSE}
library(micsPlusTableR)
run_app()
# Or select the output location:
run_app(output_dir = "survey-results")
```

Select the survey and wave, provide the plan, data files, and preparation
folder, then prepare the data. Calculate the required tables, inspect the
consistency checks, and export the workbooks. By default, workbooks are written
to `micsPlusTableR-output` below the project root detected by `here::here()`.

The **Long Format Data** tab offers Excel, CSV, and RDS downloads. Downloads
include the full prepared dataset even when the preview is filtered. RDS
preserves R column types. The SDMX-compatible fields follow the package's
observation contract; review requirements of any external SDMX system separately.

## Scripted use

A session keeps prepared data, the table plan, and results together. Create a
separate session for independent workflows:

```{r session}
library(micsPlusTableR)
session <- mics_session()
session
```

The following example requires your own input files:

```{r workflow, eval=FALSE}
prepare_mics_data(
  session,
  hh_path = "hh.sav",
  hl_path = "hl.sav",
  prep_script = "survey-preparation/survey_prep.R",
  output_dir = "survey-results"
)
result <- tabulate_mics_table(session, "TabulationPlan.xlsx", "1.1")
checks <- check_mics_table(session, result)
preview <- pivot_mics_table(session, result, type = "header")

workbooks <- create_excel_workbooks(
  "TabulationPlan.xlsx", "survey-results",
  country_code = "JAM", period = "2023-24", wave = "Wave 1"
)
write_mics_table(session, workbooks$output, sheet = "1.1", table = result)
write_mics_footnotes(session, workbooks$output, sheet = "1.1", table = result)
```

`compare_mics_tables()` and `read_previous_mics_table()` support review against
previous results. Consistency checks help find issues but cannot validate the
substantive choices in a preparation script or tabulation plan.

## Edit surveys and waves

In the development source, edit **`inst/extdata/survey_choices.csv`**. It is a
normal CSV and remains editable when stored on GitHub. R includes it in the
installed package automatically. Its current contents can be read with:

```{r survey-choices}
path <- system.file("extdata", "survey_choices.csv", package = "micsPlusTableR")
read.csv(path, colClasses = "character", fileEncoding = "UTF-8-BOM")
```

Keep the four columns `country`, `country_code`, `period`, and `wave`. Add one
row per available combination, with no blank fields or duplicates. Country
codes must be three uppercase letters and consistent within a country/period.
Treat periods such as `2025-26` as text in spreadsheet editors. Save as UTF-8.
Row order controls dropdown order; the first survey and its first wave are the
defaults. Only the selected survey's waves are offered.

From the package root, `source("run_dev.R")` loads the source files directly.
Stop the app and relaunch after saving changes. For an installed package,
reinstall the updated source and restart. Avoid maintaining the installed CSV:
installation replaces that copy.

## Documentation and distribution

Use `help(package = "micsPlusTableR")` for the function index and
`help("survey_choices", package = "micsPlusTableR")` for the CSV specification.
The detailed application guide is available at:

```{r guide}
system.file("doc", "user-guide.pdf", package = "micsPlusTableR")
```

The repository root is already a standard R source package. Commit the source
files to GitHub, including the CSV, `man/` help pages, and this vignette. From
that root, `Rscript tools/build-package.R` validates the CSV and creates an
installable source archive in `build/`. `Rscript tools/build-manual.R` creates
the CRAN-style reference PDF from the help pages in `output/pdf/`.

No conversion or duplicate package folder is needed before pushing to GitHub.
